An Extension of LTL with Rules\\ and its Application to Runtime Verification

### 2019-02-04 main.tex
Updates from ShareLaTeX

\author{ 
Klaus Havelund~\thanks{The research performed by this author was carried out at Jet Propulsion Laboratory, California Institute of Technology, under a contract with the National Aeronautics and Space Administration.} 
\and
Doron Peled~\thanks{The research performed by this author was partially funded by Israeli Science Foundation grant 2239/15: 
``Runtime Measuring and Checking of Cyber Physical Systems''.
}
}

>>>>

\author{ 
Klaus Havelund~\thanks{The research performed by this author was carried out at Jet Propulsion Laboratory, California Institute of Technology, under a contract with the National Aeronautics and Space Administration.} 
\and
Doron Peled~\thanks{The research performed by this author was partially funded by Israeli Science Foundation grant 1464/18: 
``Efficient Runtime Verification for Systems with Lots of Data and its Applications''.
}
}

### 2019-02-04 rule-ltl.tex
Updates from ShareLaTeX

\vspace{1ex}
\noindent {\bf Related Work}
For propositional LTL, the logic ETL~\cite{WVS,Wolper} extends
the set of temporal operators 
by using automata (or, equivalently a right linear grammar). Each automaton defines a temporal operator,
where the input letters of the automaton correspond to future temporal subformulas. Each accepted run defines then the places
where the subformulas (corresponding to the occurrences of the input letters in the run) need to hold. Another expressive-equivalent extension, 
in~\cite{Wolper}, adds
state-dependent quantification over propositions to LTL.

@@

The extensions we suggest are in particular easy to implement, as an extension to existing propositional/first-order runtime verification. We extend our tool \dejavu, experiment with new examples that take advantage of the extension.
\fi
\vspace{1ex}
\noindent {\bf conventions}
\label{sec:prelim}
In this paper, we treat and present several versions of the logic LTL. To simplify
the presentation, we prefix LTL with letters as follows:

\begin{itemize}
\item{P} restricts to {\em past} temporal operators.
\item{F} allows first-order quantification over
variables and relations over variables and constants. This is referred to as {\em static} quantification.
\item{Q} adds quantification over predicates (for the propositional versions)
or over relations (for the first-order version). These objects can have different values at different states. This is referred to as {\em dynamic} quantification.
\item{E} allows extending the objects that occur in the input with
new auxiliary predicates (for the propositional versions) or relations (for the first-order cases).
\end{itemize}

\klaus{You above introduce some concepts before they really are explained. Q and E. I understand why you do this (to have it in one place), but it may be too difficult for the reader to grasp if not "initiated".
Especially the E.}

\klaus{We use ELTL (etc) to denote Extended LTL. 
But QLTL is also an extension of LTL. These logics 
are all extensions of LTL. Perhaps RLTL (for Rule-based 
LTL) would have been better ... or something. Not urgent, 
we can always change that if we want to.}

@@

@@

The models for LTL formulas 
are infinite sequence of states.
$\sigma =  s_{1}\, s_{2}\, s_3 \ldots$,
where $s_i \subseteq P$ for each $i\geq 0$. These are the propositions that {\em hold} in that state.
We denote by $\xi_{i}$ the suffix $s_{i} s_{i+1} s_{i+2}\ldots$ of $\sigma$. 
LTL's semantics is defined as follows:
\begin{itemize}
\item $( \sigma, {i}) \models \mathit{true}$.
\item $(\sigma, {i}) \models p$ iff $p \in s_{i}$.
\item $(\sigma, {i}) \models \neg\varphi$ iff  $(\sigma, {i}) \not\models \varphi$.
\item $(\sigma, {i}) \models ( \varphi \land \psi)$ iff $(\sigma, {i}) \models \varphi$ and $(\sigma, {i} ) \models \psi$.
\item $(\sigma, {i}) \models \bigcirc \varphi$ iff $(\sigma, {i+1}) \models \varphi$.
\item $(\sigma, {i}) \models ( \varphi\ \cal{U}\ \psi )$ iff for some 
$j\geq i$, $(\sigma, {j}) \models \psi$,
and for all $i \leq k < j$, $(\sigma, {k}) \models \varphi$.
\item $(\sigma , i) \models \ominus \varphi$ iff $| \sigma | > 1$ and $(\sigma, i-1) \models \varphi$.
\item $(\sigma , i) \models (\varphi \since \psi)$ iff there exists $1 \leq k \leq i$ such that
$(\sigma , i) \models \psi$  and for each
$j < k \leq i$, 
$(\sigma , k)  \models \psi$.
\end{itemize}
Then $\sigma \models \varphi$ when $( \sigma ,0 ) \models \varphi$.
We can use the following abbreviations:
$\false = \neg \true$, 
$(\varphi \vee \psi) = \neg ( \neg \varphi \wedge \neg \psi )$, 
$(\varphi \rightarrow \psi) = ( \neg \varphi \vee \psi)$,
$(\varphi \leftrightarrow \psi) = ((\neg \varphi \wedge \neg \psi) \vee (\varphi \wedge \psi ))$,
$\Diamond \varphi = (\true \until \varphi)$ and $\Box \varphi = \neg \Diamond \neg \varphi$,
${\bf P}\  \varphi = (\true \; \since \; \varphi)$, and 
${\bf H}\ \varphi = \neg {\bf P}\ \neg \varphi$.

%%infinite sequences (finite executions can be extended by repeating the last event indefinitely to form an infinite paths).
The expressive power of propositional LTL is often compared to regular expressions
over the alphabet $\Sigma = 2^P$ and to  {\em monadic} logics where the temporal propositions
correspond to unary predicates (over time). A full overview appears in~\cite{Thomas}. Accordingly, we have the following characterizations:
LTL is equivalent to monadic first-order logic\footnote{First-order logic with unary predicates over the naturals, and the relation $<$.},  star-free regular expressions\footnote{Regular expressions with the following operators: constant letters, concatenation and complementation, but no star (or $\omega$-star for infinite words).} and counter-free B\"{u}chi automata\footnote{A counter-free language is a regular language for which there is an integer n such that for all words x, y, z and integers $m \geq n$ we have that $xy^m z \in L$ if and only if $xy^n z \in L$.}. In fact, restricting the temporal operators to the {\em future}  operators $\until$ and $\bigcirc$
(and the ones derived from them $\Box$ and $\Diamond$)
leaves the same expressive power and is commonly used.

@@

An important subset of LTL, called here PLTL, allows only past temporal
operators: $\since$, $\ominus$ and the ones derived
from them $\mathbf{H}$ and $\mathbf{P}$. The logic is often
(and also here) interpreted over finite sequences, 
where $\sigma \models \varphi$ when $( \sigma , | \sigma | ) \models \varphi$.
Alternatively,
it is also a common practice to prefix the property with a single $\Box$ (always) operator, where each of the prefixes has to satisfy $\varphi$. This later
form expresses {\em safety} properties~\cite{AS} over infinite sequences.

@@

\section*{Extending Propositional LTL}
\label{sec:extending-prop-ltl}

\subsubsection{Extending with Dynamic Quantification}

Wolper~\cite{Wolper} showed that the expressive power of LTL is lacking with respect
to some important properties. As a canonical example, he demonstrated that it is impossible to
express that all the states with even indexes of a sequence
satisfy $p$ (this is
different than saying that $p$ alternates between $\true$ and $\false$ on consecutive states). In order to extend the expressiveness of LTL, Wolper suggested to use linear grammars, or, alternatively, the ability to quantify over
propositions.

Adding quantification over propositions allows writing a formula of the form
$\Exists q \, \varphi$, where $q$ is a proposition that can appear in $\varphi$.
Let the propositions $P$ be partitioned into two disjoint sets: 
{\em auxiliary propositions} $A = \{ a_1 , \; \ldots\; , a_n \}$
and {\em basic propositions} $B$. 
%A model for such a formula 
%has truth assignments only over the propositions in $B$, but the propositions $A$ can appear (only) within the scope of a quantifier.
We also allow $\Forall q\, \varphi =
\neg \Exists q\, \neg \varphi$.
To define the semantics, let $X$ be a subset of 
the propositional variables appearing in the states of the sequence $\sigma = s_1 s_2 \ldots$ and 
denote $\sigma |_X = s_1 \setminus X \, s_2 \setminus X \ldots$. (Note that $\sigma |_X$ denotes projecting {\em out} the variables in $X$.)

\begin{itemize}
\item $( \sigma, i ) \models \boldsymbol{\Exists} q \, \varphi$ iff there exists
$\sigma'$ such that $\sigma' |_{\{ q \} } = \sigma$ and 
$( \sigma', i) \models \varphi$.
\end{itemize}
Since all the auxiliary variables appear in the scope of
quantifiers, it holds that $\sigma \models \varphi$
(in $\varphi$, auxiliary variables appear only within
the scope of quantifiers)
iff $\sigma |_A \models \varphi$.

\klaus{Are the last two lines above needed?}

This kind of quantification is considered to be ``dynamic'', since the quantified variables can have different truth values depending on the states, or, "second order", the quantification establishes the {\em set} of states in which a propositional variable has the value $\true$.
Extending LTL with such quantification, the logic QLTL has the same expressive power as regular expressions, full B\"{u}chi
automata, or monadic second order logic with unary predicates over the naturals, see again~\cite{Thomas}.
In fact, it is sufficient to restrict the quantification to existential quantifiers that prefix the formula to obtain the full expressiveness of QLTL~\cite{Thomas}.

\noindent
Wolper's property can be expressed in QLTL as follows:
\begin{equation} \label{form0} \Exists q \,
( (  \neg q \wedge \Box (q \leftrightarrow \bigcirc \neg q )  )  \wedge \Box (q \rightarrow p ) )\end{equation}
Wolper's example property cannot be expressed in PLTL either.

@@

\subsubsection{Extending with Rules}

@@

\item $( \gamma , {\cal R} , i ) \models ( \varphi \; \until \; \psi )$ if
for some $j > i$, $( \gamma |_{\vars ( \psi )} , {\cal R} , j) 
\models \psi $ and for all $i \leq k < j $,
$( \gamma |_{\vars ( \varphi )} , {\cal R} , k) \models \varphi$.

@@

\item $( \gamma , {\cal R} , i ) \models ( \varphi \; \since \; \psi )$ if
for some $1 \le j \leq i$, $( \gamma |_{\vars ( \psi )} , {\cal R} , j) 
\models \psi $ and for all $j < k \leq i$,
$( \gamma |_{\vars ( \varphi )} , {\cal R} , k) \models \varphi$.

@@

\section*{Extending First-Order LTL}

@@

\subsubsection{Extending with Quantification}

@@

\subsubsection{Extending with Rules}

>>>>

\vspace{1ex}
\noindent {\bf Related Work.}
For propositional LTL, the logic ETL~\cite{WVS,Wolper} extends
the set of temporal operators 
by using automata (or, equivalently a right linear grammar). Each automaton defines a temporal operator,
where the input letters of the automaton correspond to future temporal subformulas. Each accepted run defines then the places
where the subformulas (corresponding to the occurrences of the input letters in the run) need to hold. Another expressive-equivalent extension, 
in~\cite{Wolper}, adds
state-dependent quantification over propositions to LTL.

@@

The extensions we suggest are in particular easy to implement, as an extension to existing propositional/first-order runtime verification. We extend our tool \dejavu, experiment with new examples that take advantage of the extension.
\fi
\vspace{1ex}
\noindent {\bf Conventions.}
\label{sec:prelim}
In this paper, we treat and present several versions of the logic LTL. To simplify
the presentation, we prefix LTL with letters as follows:

\begin{itemize}
\item{P} restricts to P{\em ast} temporal operators.
\item{F} allows F{\em irst-order} quantification over
data assigned to variables that is uniform for all states. This quantification is referred to as {\em static} quantification.
\item{Q} adds (second order) Q{\em uantification} that is state-dependent. This is referred to as {\em dynamic} quantification.
\item{E} allows E{\em xtending} the objects that occur in the input with
new auxiliary predicates (for the propositional versions) or relations (for the first-order cases) and past temporal logic rules that define them.
\end{itemize}

%\klaus{You above introduce some concepts before they really are explained. Q and E. I understand why you do this (to have it in one place), but it may be too difficult for the reader to grasp if not "initiated".
%Especially the E.}

%\klaus{We use ELTL (etc) to denote Extended LTL. 
%But QLTL is also an extension of LTL. These logics 
%are all extensions of LTL. Perhaps RLTL (for Rule-based 
%LTL) would have been better ... or something. Not urgent, 
%we can always change that if we want to.}

@@

\doron{Did my best to make the letters more self explained. I think R is usually reserved in temporal logic for realtime. But if you insist, one thing that I cannot perform from here is changing the two figures where these letters occur.}

@@

The models for LTL formulas 
are infinite sequence of states, of the form
$\sigma =  s_{1}\, s_{2}\, s_3 \ldots$,
where $s_i \subseteq P$ for each $i\geq 0$. These are the propositions that {\em hold} in that state.
We denote by $\xi_{i}$ the suffix $s_{i} s_{i+1} s_{i+2}\ldots$ of $\sigma$. 
LTL's semantics is defined as follows:
\begin{itemize}
\item $( \sigma, {i}) \models \mathit{true}$.
\item $(\sigma, {i}) \models p$ iff $p \in s_{i}$.
\item $(\sigma, {i}) \models \neg\varphi$ iff  $(\sigma, {i}) \not\models \varphi$.
\item $(\sigma, {i}) \models ( \varphi \land \psi)$ iff $(\sigma, {i}) \models \varphi$ and $(\sigma, {i} ) \models \psi$.
\item $(\sigma, {i}) \models \bigcirc \varphi$ iff $(\sigma, {i+1}) \models \varphi$.
\item $(\sigma, {i}) \models ( \varphi \cal{U} \psi )$ iff for some $j$,
$j\geq i$, $(\sigma, {j}) \models \psi$,
and for each $k$, $i \leq k < j$, $(\sigma, {k}) \models \varphi$.
\item $(\sigma , i) \models \ominus \varphi$ iff $| \sigma | > 1$ and $(\sigma, i-1) \models \varphi$.
\item $(\sigma , i) \models (\varphi \since \psi)$ iff there exists $j$, $1 \leq j \leq i$, such that
$(\sigma , j)  \models \psi$ and for each
$k$, $j < k \leq i$, $(\sigma , k ) \models \varphi$.
\end{itemize}
Then $\sigma \models \varphi$ when $( \sigma ,0 ) \models \varphi$.
We can use the following abbreviations:
$\false = \neg \true$, 
$(\varphi \vee \psi) = \neg ( \neg \varphi \wedge \neg \psi )$, 
$(\varphi \rightarrow \psi) = ( \neg \varphi \vee \psi)$,
$(\varphi \leftrightarrow \psi) = ((\neg \varphi \wedge \neg \psi) \vee (\varphi \wedge \psi ))$,
$\Diamond \varphi = (\true \until \varphi)$ and $\Box \varphi = \neg \Diamond \neg \varphi$,
${\bf P}\  \varphi = (\true \; \since \; \varphi)$ ({\bf P} stands for P{\em reviously}), and 
${\bf H}\  \varphi = \neg {\bf P}\ \neg \varphi$.

%%infinite sequences (finite executions can be extended by repeating the last event indefinitely to form an infinite paths).
The expressive power of propositional LTL is often compared to regular expressions
over the alphabet $\Sigma = 2^P$ and to  {\em monadic} first and second order logics where the temporal propositions
correspond to unary predicates (over time). An overview appears in~\cite{Thomas}. Accordingly, we have the following characterizations:
LTL is equivalent to monadic first-order logic\footnote{First-order logic with unary predicates over the naturals, and the relation $<$.},  star-free regular expressions\footnote{Regular expressions but without star (or $\omega$-star for infinite words).} and counter-free B\"{u}chi automata\footnote{A counter-free language is a regular language for which there is an integer $n$ such that for all words x, y, z and integers $m \geq n$ we have that $xy^m z \in L$ if and only if $xy^n z \in L$.}. In fact, restricting the temporal operators to the {\em future}  operators $\until$ and $\bigcirc$
(and the ones derived from them $\Box$ and $\Diamond$)
leaves the same expressive power and is commonly used.

@@

An important subset of LTL, called here PLTL, allows only past temporal
operators: $\since$, $\ominus$ and the ones derived
from them, $\mathbf{H}$ and $\mathbf{P}$. The past time logic is often
(and also here) interpreted over finite sequences, 
where $\sigma \models \varphi$ when $( \sigma , | \sigma | ) \models \varphi$.
Alternatively,
it is also a common practice to use the past time restricted version, prefixed with  a single $\Box$ (always) operator; there, each of the prefixes has to satisfy $\varphi$. This later
form expresses {\em safety} LTL properties~\cite{AS} over infinite sequences.

@@

%\section*{Extending Propositional LTL}
\label{sec:extending-prop-ltl}

Wolper~\cite{Wolper} showed that the expressive power of LTL is lacking with respect
to some important properties. As a canonical example, he demonstrated that it is impossible to
express that all the states with even indexes in a sequence
satisfy $p$ (this is
different than saying that $p$ alternates between $\true$ and $\false$ on consecutive states).
Wolper's example property cannot be expressed in PLTL either.
In order to extend the expressiveness of LTL, Wolper suggested to use linear grammars, or, alternatively, the ability to quantify over
propositions, as shown below.

\subsubsection{Extending LTL with Dynamic Quantification.}
Adding quantification over propositions allows writing a formula of the form
$\Exists q \, \varphi$, where
$\Exists q$ repesents {\em existential} quantification over a propositiona $q$ that can appear in $\varphi$.
Let the propositions $P$ be partitioned into two disjoint sets: 
{\em auxiliary propositions} $A = \{ a_1 , \; \ldots\; , a_n \}$
and {\em basic propositions} $B$. 
%A model for such a formula 
%has truth assignments only over the propositions in $B$, but the propositions $A$ can appear (only) within the scope of a quantifier.

To define the semantics, let $X \subseteq P$ 
%be a subset of 
%the propositional variables appearing in the %states of the sequence $\sigma = s_1 s_2 %\ldots$ 
and denote $\sigma |_X = s_1 \setminus X \, s_2 \setminus X \ldots$. (Note that $\sigma |_X$ denotes projecting {\em out} the variables in $X$.)

\begin{itemize}
\item $( \sigma, i ) \models \boldsymbol{\Exists} q \, \varphi$ iff there exists
$\sigma'$ such that $\sigma' |_{\{ q \} } = \sigma$ and 
$( \sigma', i) \models \varphi$.
\end{itemize}
%Since all the auxiliary variables appear in the scope of
%quantifiers, it holds that $\sigma \models \varphi$
%(in $\varphi$, auxiliary variables appear only within
%the scope of quantifiers)
%iff $\sigma |_A \models \varphi$.
%
%\klaus{Are the last two lines above needed?}
%
Universal quantification is also allowed, where $\Forall q\, \varphi =
\neg \Exists q\, \neg \varphi$.
This kind of quantification is considered to be ``dynamic'', since the quantified variables can have different truth values depending on the states It is also callsed  "second order" quantification, since the quantification establishes the {\em set} of states in which a propositional variable has the value $\true$.
Extending LTL with such quantification, the logic QLTL has the same expressive power as regular expressions (over infinite words), full B\"{u}chi
automata, or monadic second order logic with unary predicates over the naturals, see again~\cite{Thomas}.
In fact, it is sufficient to restrict the quantification to existential quantifiers that prefix the formula to obtain the full expressiveness of QLTL~\cite{Thomas}.

\noindent
Wolper's property can be expressed in QLTL as follows:
\begin{equation} \label{form0} \Exists q \,
( (  \neg q \wedge \Box (q \leftrightarrow \bigcirc \neg q )  )  \wedge \Box (q \rightarrow p ) )\end{equation}

@@

\subsubsection{Extending LTL with Rules.}

@@

\item $( \gamma , {\cal R} , i ) \models ( \varphi \; \until \; \psi )$ if
for some $j$, $j > i$, $( \gamma |_{\vars ( \psi )} , {\cal R} , j) 
\models \psi $ and for each $k$,  $i \leq k < j $,
$( \gamma |_{\vars ( \varphi )} , {\cal R} , k) \models \varphi$.

@@

\item $( \gamma , {\cal R} , i ) \models ( \varphi \; \since \; \psi )$ if
for some $j$, $1 \le j \leq i$, $( \gamma |_{\vars ( \psi )} , {\cal R} , j) 
\models \psi $ and for each $k$, $j < k \leq i$,
$( \gamma |_{\vars ( \varphi )} , {\cal R} , k) \models \varphi$.

@@

%\section*{Extending First-Order LTL}

@@

\vspace{1ex}
\noindent {\bf Example 2.}
Consider the property that asserts that
a main process $y$ (where
$mathit{main} (x)$ appears
in the sequence), is a descendant of a process $x$ if $x$ spawns a chain of processes spawning (of
the form $\mathit{spawn} (z, w$ that leadsfrom $x$ to $y$. This property cannot be expressed in FLTL. The reason
is that the required property needs to
assert about the transitive closure
of the relation $\mathit{spawn}$. 
FLTL can be translated in a way similar to a standard translation of $LTL$ to monadic first order logic~\cite{Thomas}. Here, the relations will be written with their explicit last ``time'' parameter, and the temporal operators are replaced with first order quantification.
E.g., $\Box \forall x \, p(x) \rightarrow q(y)$ will be translated into $\forall t \, \forall x \, (p(x, i) \rightarrow \exists t' p(x, j) )$
\footnote{The obtain model will have, in addition
to the aforementioned relations, also
the linear order $<$ over the naturals.}
It is possible to show that the transitive closure of $\mathit{spawn}$ cannot be expressed in this first order setting. The proof is based on the
compactness theory of first order logic~\cite{Flum}.

\subsubsection{Extending FLTL with Quantification.}

@@

\subsubsection{ExtendingFLTL  with Rules.}
